---
title: "Homework 1"
author: Marisa Blackman
date: January 15, 2023
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Code from mixture-data-li-knn.R with edits for assignment

## Data

```{r}
library('class')
library('dplyr')

## load binary classification example data from author website 
## 'ElemStatLearn' package no longer available
load(url('https://web.stanford.edu/~hastie/ElemStatLearn/datasets/ESL.mixture.rda'))
dat <- ESL.mixture

plot_mix_data <- function(dat, datboot=NULL) {
  if(!is.null(datboot)) {
    dat$x <- datboot$x
    dat$y <- datboot$y
  }
  plot(dat$x[,1], dat$x[,2],
       col=ifelse(dat$y==0, 'blue', 'orange'),
       pch=20,
       xlab=expression(x[1]),
       ylab=expression(x[2]))
  ## draw Bayes (True) classification boundary
  prob <- matrix(dat$prob, length(dat$px1), length(dat$px2))
  cont <- contourLines(dat$px1, dat$px2, prob, levels=0.5)
  rslt <- sapply(cont, lines, col='purple')
}

plot_mix_data(dat)
```

## Linear Classifier with lm and predict functions

```{r}
## fit linear classifier
fit_lc <- function(y, x1, x2, data) {lm(y ~ x1 + x2, data = data)}

## make predictions from linear classifier
predict_lc <- function(fit, newx) {predict(fit, newx)}

## fit model to mixture data and make predictions #on grid of values
lc_lm <- fit_lc(dat$y, dat$x[,1], dat$x[,2], data.frame(dat$y, dat$x)) 
lc_pred <- predict_lc(lc_lm, dat$xnew)  

## reshape predictions as a matrix
lc_pred <- matrix(lc_pred, length(dat$px1), length(dat$px2))
contour(lc_pred,
        xlab=expression(x[1]),
        ylab=expression(x[2]))

## find the contours in 2D space such that lc_pred == 0.5
lc_cont <- contourLines(dat$px1, dat$px2, lc_pred, levels=0.5)

## plot data and decision surface
plot_mix_data(dat)
sapply(lc_cont, lines)
```

## Linear classifier with squared terms

```{r}
## fit linear classifier
fit_lc <- function(y, x1, x2, data) {lm(y ~ x1 + x2 + I(x1^2) + I(x2^2), data = data)}

## make predictions from linear classifier
predict_lc <- function(fit, newx) {predict(fit, newx)}

## fit model to mixture data and make predictions #on grid of values
lc_lm <- fit_lc(dat$y, dat$x[,1], dat$x[,2], data.frame(dat$y, dat$x)) 
lc_pred <- predict_lc(lc_lm, dat$xnew)  

## reshape predictions as a matrix
lc_pred <- matrix(lc_pred, length(dat$px1), length(dat$px2))
contour(lc_pred,
        xlab=expression(x[1]),
        ylab=expression(x[2]))

## find the contours in 2D space such that lc_pred == 0.5
lc_cont_squared <- contourLines(dat$px1, dat$px2, lc_pred, levels=0.5)

## plot data and decision surface
plot_mix_data(dat)
sapply(lc_cont, lines, col = "black")
sapply(lc_cont_squared, lines, col = "red")
legend(-2, -1, legend=c("Linear", "Squared terms"),
       col=c("black", "red"), lty=1, cex=0.8)
```

## Describe how this more flexible model affects the bias-variance tradeoff

Based on the above graph, the more flexible model has more bias since it is more fit to our specific data. This also reduces variance since the distance between observations and predictions should be smaller. 

## KNN Classifier

```{r}
## fit knn classifier
## use 5-NN to estimate probability of class assignment
knn_fit <- knn(train=dat$x, test=dat$xnew, cl=dat$y, k=5, prob=TRUE)
knn_pred <- attr(knn_fit, 'prob')
knn_pred <- ifelse(knn_fit == 1, knn_pred, 1-knn_pred)

## reshape predictions as a matrix
knn_pred <- matrix(knn_pred, length(dat$px1), length(dat$px2))
contour(knn_pred,
        xlab=expression(x[1]),
        ylab=expression(x[2]),
        levels=c(0.25, 0.5, 0.75))


## find the contours in 2D space such that knn_pred == 0.5
knn_cont <- contourLines(dat$px1, dat$px2, knn_pred, levels=0.5)

## plot data and decision surface
plot_mix_data(dat)
sapply(knn_cont, lines)


## do bootstrap to get a sense of variance in decision surface
resample <- function(dat) {
  idx <- sample(1:length(dat$y), replace = T)
  dat$y <- dat$y[idx]
  dat$x <- dat$x[idx,]
  return(dat)
}

## plot linear classifier for three bootstraps
par(mfrow=c(1,3))
for(b in 1:3) {
  datb <- resample(dat)
  ## fit model to mixture data and make predictions
  lc_beta <- fit_lc(datb$y, datb$x)
  lc_pred <- predict_lc(datb$xnew, lc_beta)
  
  ## reshape predictions as a matrix
  lc_pred <- matrix(lc_pred, length(datb$px1), length(datb$px2))
  
  ## find the contours in 2D space such that lc_pred == 0.5
  lc_cont <- contourLines(datb$px1, datb$px2, lc_pred, levels=0.5)
  
  ## plot data and decision surface
  plot_mix_data(dat, datb)
  sapply(lc_cont, lines)
}

## plot 5-NN classifier for three bootstraps
par(mfrow=c(1,3))
for(b in 1:3) {
  datb <- resample(dat)
  
  knn_fit <- knn(train=datb$x, test=datb$xnew, cl=datb$y, k=5, prob=TRUE)
  knn_pred <- attr(knn_fit, 'prob')
  knn_pred <- ifelse(knn_fit == 1, knn_pred, 1-knn_pred)
  
  ## reshape predictions as a matrix
  knn_pred <- matrix(knn_pred, length(datb$px1), length(datb$px2))
  
  ## find the contours in 2D space such that knn_pred == 0.5
  knn_cont <- contourLines(datb$px1, datb$px2, knn_pred, levels=0.5)
  
  ## plot data and decision surface
  plot_mix_data(dat, datb)
  sapply(knn_cont, lines)
}

## plot 20-NN classifier for three bootstraps
par(mfrow=c(1,3))
for(b in 1:3) {
  datb <- resample(dat)
  
  knn_fit <- knn(train=datb$x, test=datb$xnew, cl=datb$y, k=20, prob=TRUE)
  knn_pred <- attr(knn_fit, 'prob')
  knn_pred <- ifelse(knn_fit == 1, knn_pred, 1-knn_pred)
  
  ## reshape predictions as a matrix
  knn_pred <- matrix(knn_pred, length(datb$px1), length(datb$px2))
  
  ## find the contours in 2D space such that knn_pred == 0.5
  knn_cont <- contourLines(datb$px1, datb$px2, knn_pred, levels=0.5)
  
  ## plot data and decision surface
  plot_mix_data(dat, datb)
  sapply(knn_cont, lines)
}
```

